---
title: "Hotel Bookings EDA"
author: "Sergio Puno"
date: "2/23/2022"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: paper
    collapsed: true
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Preparations {.tabset}
Below are the required packages used if you'd like to reproduce the results. The full
source code is available at [my github repository](https://github.com/Sergio-Puno).

## Load libraries
Libraries for data wrangling, general visualization and extra specialized tools.
```{r}
# data manipulation
library(tidyverse)
library(tidytuesdayR)
library(lubridate)
library(tidytext)
library(rlang)

# visualization
library(ggplot2)
library(scales)

# forecast

# formatting
library(kableExtra)

```

## Helper functions
First create functions that can help with some plotting to reduce repeated code chunks.
```{r}
# quick histogram plotting with median feed-in
plot_distribution <- function(df, feature, x_intercept) {
  x <- ggplot(data = df, aes(x = {{feature}})) +
    geom_histogram(fill = "steelblue3") +
    geom_vline(xintercept = as.double(x_intercept), color = "red")
  
  return(x)
}

# quick bar plot for univariate analysis
plot_bar <- function(df, feature, agg_n) {
  x <- ggplot(data = df, aes(x = {{agg_n}}, y = {{feature}}, fill = {{feature}})) +
    geom_col() +
    theme(legend.position="none")
  
  return(x)
}
```

## Load data
Import file from TidyTuesday repo and local country abbreviations cross reference csv.
```{r}
hotels <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv')

country_codes <- read_csv("country_codes.csv")

head(hotels)
```

# Data Background
This information comes from two datasets of hotel demand data, one a resort hotel
in the region of Algarve and the other a city hotel in the region of Lisbon (both in Portugal).

The brief associated with this data can be found [here](https://www.sciencedirect.com/science/article/pii/S2352340918315191#f0010).
One consideration mentioned in this briefing was the below:
<br />

*"...even though the capture of data took considered a timespan prior to arrival date, it is understandable that the distribution of some variables differ between non canceled and canceled bookings."*
<br />

We'll have to consider the impact of a cancelled booking on each distribution and how
a cancelled booking is different (on average) compared to a booking that showed up.

# Data Cleaning / Prelimary Analysis
## Missing Values
Quick missing values check to see if there are any massive holes in the data:

```{r}
# checking for missing values
missing_values <- hotels %>% 
  summarise(across(everything(), ~sum(is.na(.))))

missing_values <- gather(missing_values, key = "feature", value = "missing_pct")

missing_values %>%
  ggplot(aes(x = reorder(feature, missing_pct), y = missing_pct)) +
  geom_bar(stat = "identity", fill = "tomato2") +
  coord_flip() + 
  theme_bw()
```

<br />

Looks clean, only the children records seem to have missing values, about 4% of the records
are missing child data, this is not a huge deal so we can keep these records.

### Creating/Updating data fields
Generating any new fields/columns that I feel may be helpful in subsequent analysis
as well as changing some data types from the initial import such as converting certain
fields to factors.

```{r}
# creating a new date column with the year/month/day of month cols for arrivals
hotels <- hotels %>% 
  mutate(arrival_date = dmy(paste(arrival_date_day_of_month,
                                  arrival_date_month,
                                  arrival_date_year,
                                  sep = "-"))) %>% 
  mutate(is_canceled = as.factor(is_canceled),
         stays_in_weekend_nights = as.factor(stays_in_weekend_nights),
         stays_in_week_nights = as.factor(stays_in_week_nights),
         meal = as.factor(meal),
         market_segment = as.factor(market_segment),
         distribution_channel = as.factor(distribution_channel),
         reserved_room_type = as.factor(reserved_room_type),
         assigned_room_type = as.factor(assigned_room_type),
         deposit_type = as.factor(deposit_type),
         customer_type = as.factor(customer_type),
         reservation_status = as.factor(reservation_status))
```

## Univariate Analysis {.tabset}
Distributions and getting to know our data

### Market Segments
Checking most common segments that drive bookings
```{r}
hotels %>% 
  filter(market_segment != "Undefined") %>% 
  count(market_segment, sort= TRUE) %>% 
  plot_bar(., market_segment, n) +
  labs(x = "# of Bookings", y = "Market Segment",
       title = "Number of Bookings by Market Segment") +
  scale_x_continuous(labels = comma)

hotels %>% 
  count(market_segment, sort = TRUE)
```

### Countries of visitors
Identifying most common customer country of origin, bringing in full country name from
the helper file I loaded in for country abbreviations.
```{r}
hotels %>% 
  left_join(., country_codes, by = "country") %>% 
  filter(!is.na(country_name)) %>% 
  mutate(country_name = fct_lump(country_name, 20)) %>% 
  count(country_name, sort= TRUE) %>% 
  mutate(country_name = fct_reorder(country_name, n)) %>% 
  plot_bar(., country_name, n) +
  labs(x = "# of Bookings", y = "Country of Origin",
       title = "Number of Bookings by Country") +
  scale_x_continuous(labels = comma)

hotels %>% 
  count(country, sort= TRUE)
```

### Customer Type
What is the most common customer type, we can study this in combination with other
details we have such as `reserved_room_type` or `deposit_Type` to see if certain
types of customers are aligned with certain other requirements.
```{r}
hotels %>% 
  count(customer_type, sort = TRUE) %>% 
  plot_bar(., customer_type, n) +
  labs(x = "Customer Type", y = "",
       title = "Customer Types")

hotels %>% 
  count(customer_type, sort = TRUE)
```

### ADR
Average Daily Rate - this is the only value field we have related to cost in this dataset, with this we 
can take a look later on at how where some of the rate is derived from (such as the room type, 
market segment, meal type, etc.)

```{r}
adr_median <- hotels %>% 
  filter(adr > 0) %>% 
  summarise(median = median(adr))

hotels %>% 
  filter(adr > 0) %>% 
  plot_distribution(., adr, adr_median) +
  scale_x_log10(labels = dollar)

# top 10 ADR values
hotels %>% 
  arrange(desc(adr)) %>% 
  head(10) %>% 
  select(hotel, is_canceled, arrival_date, assigned_room_type, adr)
```

<br />

# Market Segment Analysis and Impact

### Market segment breakdown with overall cancellations by hotel type
Checking out the top market segments of each hotel.

```{r}
# data manipulation for plotting
hotels_markets <- hotels %>% 
  filter(market_segment != "Undefined") %>% 
  mutate(market_segment = fct_lump(market_segment, 4)) %>%
  count(hotel, market_segment, is_canceled) %>% 
  group_by(hotel) %>% 
  mutate(total = sum(n),
         ratio = n / total,
         is_canceled = ifelse(is_canceled == 0, "No", "Yes")) %>% 
  ungroup() %>% 
  mutate(market_segment = reorder_within(market_segment, ratio, hotel))

hotels_markets %>% 
  ggplot(aes(ratio, market_segment, fill = is_canceled)) +
  geom_col(position = "dodge") +
  labs(x = "Percentage of Bookings", y = "Market Segment",
       title = "Hotel Bookings by Market Segment: Hotel Split",
       subtitle = "Sorted top-down, most used to least used",
       fill = "Booking Canceled") +
  facet_wrap(~ hotel, scales = "free_y") +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("steelblue2", "tomato2")) +
  scale_y_reordered() +
  theme(legend.position = "top")

```

<br />

Looking at the results, we can see that the majority of people use travel agents
(both online and offline) with direct being a close follow up for the resort hotel.
We also can see that cancellations are quite proportionally high for the city hotel, 
especially for the group bookings which makes sense as there is a high chance for one
individual out of a group to cause a cancellation due to external factors.


```{r}
hotels_markets <- hotels %>% 
  filter(market_segment != "Undefined") %>% 
  group_by(market_segment) %>% 
  count(market_segment, is_canceled) %>%
  mutate(total = sum(n),
         ratio = n / total,
         is_canceled = ifelse(is_canceled == 0, "No", "Yes")) %>% 
  arrange(desc(n))

# stacked column chart grouping all bookings
hotels_markets %>% 
  ggplot(aes(market_segment, ratio, fill = is_canceled)) +
  geom_col() +
  labs(x = "Market Segment", y = "Percentage of Bookings",
     title = "Hotel Bookings by Market Segment: Combined",
     fill = "Booking Canceled") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_reordered() +
  scale_fill_manual(values = c("steelblue2", "tomato2")) +
  theme(legend.position = "top") +
  coord_flip()
```

<br />

We can see from this stacked bar chart that overall, for bookings using online 
travel agents there are between 15% to almost 20% cancellations. Interestingly, 
compared to the online agents, the `Direct` and `Complementary` bookings have the
lowest cancellation rates with `Corporate` following closely behind. 

<br />

# Lead Time Analysis
Taking a look at the `lead_time` values, we can split the lead times into two 
categories, same day bookings and standard ahead of date bookings.

A few things I have in mind while reviewing the distributions, many of these apply
to all distributions I'll look at and not just lead-time specific:
<ul>
<li> How extreme are outliers compared to the majority of the data's distribution
<li> Check for any bimodal distributions or any cycles
<li> Would the cancellation vs. non-cancellation factor impact the distributions
</ul>
```{r}
# median for non 0 LT (lead time) bookings
nonzero_LT_median <- hotels %>% 
  filter(lead_time > 0) %>% 
  summarise(median = median(lead_time))

# for non same-day bookings, was does the lead time distribution look like
hotels %>% 
  filter(lead_time > 0) %>% 
  plot_distribution(., lead_time, nonzero_LT_median) +
  labs(x = "Booking Lead Time", y = "Count",
       title = "Lead Time Distribution: non same day bookings")

```

### Arrival date analysis:
Looking at the breakdown of arrival day-of-week by market segment first.

```{r}
# looking at the different market segments and arrival day of week for comparisons

hotels %>% 
  mutate(wday = wday(arrival_date, label = TRUE, week_start = 1)) %>%
  filter(market_segment != "Undefined") %>% 
  group_by(wday, market_segment) %>% 
  count(wday) %>% 
  ggplot(aes(market_segment, y= n, color= wday)) +
  geom_point(size=4) +
  theme(legend.position = "left",
        plot.title = element_text(size=14)) +
  coord_flip() +
  scale_x_discrete(position="to") +
  labs(x = "") +
  ggtitle("Arrival by Weekday") +
  scale_color_hue() +
  scale_y_log10()
```

<br />

Looking at the results we find:
<ul>
<li> `Online TA`, `Offline TA/TO`, `Groups` appear to be the top three markets;
TA = Travel Agent, TO = Tour Operators
<li> You can see that the most common day of week shifts from segment to segment
with Friday being popular for `Online TA` and `Offline TA/TO` but Monday-Wednesday
being more common for `Corporate` which makes sense as if you are working you'll
likely be staying during just the work week
</ul>

<br />

### Breakdown by day of month
We can see that the first week fo the month is universally lower in each market
when compared to say the end of the month which is the top count for all markets 
except for `Aviation` which has the third week as their peak.
```{r}
hotels %>% 
  mutate(month_week = case_when(arrival_date_day_of_month >= 21 ~ '4th WK',
                                arrival_date_day_of_month >= 14 ~ '3rd WK',
                                arrival_date_day_of_month >= 7 ~ '2nd WK',
                                arrival_date_day_of_month < 7 ~ '1st WK')) %>% 
  filter(market_segment != "Undefined") %>%
  group_by(month_week, market_segment) %>% 
  count(month_week) %>% 
  ggplot(aes(market_segment, y= n, color= month_week)) +
  geom_point(size=4) +
  theme(legend.position = "left",
        plot.title = element_text(size=14)) +
  coord_flip() +
  scale_x_discrete(position="to") +
  labs(x = "") +
  ggtitle("Arrival by Week of Month") +
  scale_color_hue() +
  scale_y_log10()
```

<br />

Nothing too spread out for this plot, the main consistency is that the last week
of the month is almost always the number one week, with `Aviation` having the
second week as their most common. We can change this up a bit to see if any months
in particular show trends of bookings.
```{r}
hotels %>% 
  filter(between(arrival_date, as.Date("2016-01-01"), as.Date("2016-12-31"))) %>% 
  group_by(hotel, arrival_date, arrival_date_year) %>% 
  summarise(count = n()) %>% 
  ggplot(aes(arrival_date, count)) +
  geom_line(aes(color = hotel), size = 1) +
  facet_wrap(~ hotel, ncol = 1) +
  theme(legend.position = "none") +
    labs(x = "Arrival Date", y = "# of Bookings",
       title = "Number of Bookings (2016)")

hotels %>% 
  group_by(hotel, arrival_date_week_number) %>% 
  summarise(avg_bookings = mean(n())) %>% 
  mutate(year_month = ymd("2000-01-01") + weeks(arrival_date_week_number - 1)) %>% 
  ggplot(aes(year_month, avg_bookings)) +
  geom_line(aes(color = hotel), size = 2) +
  scale_x_date(date_labels = "%b-%d") +
  facet_wrap(~ hotel, ncol = 1, scales = "free_y") +
  theme(legend.position = "none") +
  labs(x = "Arrival Date", y = "Avg Bookings",
       title = "Average bookings: 2015-07-01 to 2017-8-31")
  
```

